---
title: |
  | NYU SURP 2019 R course
  | Class 3: Functions and Vectors
author:
- Eugene Plavskin
- Grace Avecilla
date: "06/27/2019"
output:
  html_notebook:
    code_folding: show
    depth: 3
    tidy: yes
    toc: yes
---
# Review from last time

Let's review variables and types

```{r}
# Create a variable that holds the number 53

# Create a variable that holds two times the value of the first variable

# Create a variable that holds the word "biology"

# Get R to tell you the "type" of the last variable you created

# What happens if you try to add the two "biology" variables together? Why?

```

# Functions
## Adding with functions
We've seen that you can do basic arithmetic in R by just using mathematical symbols, e.g. __+__. But for more complicated things, programming languages use things called *functions*. A function is something that wraps a whole set of operations so that you can invoke those operations in a single command. You have already used another function in the last class, `is()`. Let's take a look at another example: there's a function, `sum()`, that can add numbers, just like __+__ can. Let's try it:
```{r}
variable1 <- 5
variable2 <- 3
sum(variable1, variable2)
```

Functions are their own type! You can try using `is()` to check this:
```{r}
is(sum)
is(is)
```

Functions are always followed by `()` when you use them. When you use a function, often you are telling it to perform an operation on something else (e.g. a variable); in the examples above, we're performing the `sum()` function on *variable1* and *variable2*, and we're performing the `is()` function on `sum` and on itself. These are called arguments, and are *passed* to the function inside the parentheses.

## Looking at the **Environment**
Some functions don't need to take any arguments. For example, you can find out what variables you have loaded in your *environment* at anytime using `ls()`. This will list all your variables, any custom functions you may have written, etc.
```{r}
ls()
```

Check out the **Environment** window on the top right! You can actually also see all your variables loaded in there, with a short summary of what's in them. This is a super convenient feature of programming in Rstudio.

## Combining strings
Another really useful function is the `paste` function; this function allows you to combine *strings*.
```{r}
research_program <- 'SURP'
year <- '2019'

current_program <- paste(research_program, year)
```

Notice that the code above didn't print anything out. That's because we directed the output of the `paste` function into a variable, *current_program*. To get R to print out the current program variable, you could just type it into your console, or you could use a function! 

***
*Skip if behind*: 
Try googling to find an R function that prints out a variable.
```{r}
# Find an R function to print out a variable, and use it to print whatever
# the current_program variable is storing
```
***

You can learn about a function by reading its documentation. You can access the documentation with `?`, for example, type `?paste` into your console. You will see the help text pop up in the bottom right Help box in Rstudio.

`?` is one of the more important keys for you to know. Seriously, many issues are averted by first checking the documentation. Documentation usually has: 

* A description of the intention of the function
* How to use it, ie where to put what arguments
* A description of what those arguments should be
* Details, and what value they return
* And if you're lucky, examples of usage

In practice, this information can sometimes be a bit dense and technical. It can be sometimes helpful to just read the 'description' on top, and then scroll all the way down to the 'examples' section at the bottom of the help function.

Notice that when we used `paste` above, the function automatically put a space between our two strings. It's often really useful to combine strings using a different character between them, for example, a dash. Try to read the documentation for `paste()` to figure out how to do this.
```{r}
# Combine the research_program and year variables using just a dash between
# them instead of a space

```

*** 
*Skip if behind*
Finally, paste and sum can actually both take more than two inputs!
```{r}
# Try using sum to get the sum of 3, 4, and 5

# Try using paste to combine the research program and year variables, as well as
# a string that says "NYU", separating the 3 strings with a period

```
***

# Introduction to vectors

So far, we've been using R to do things we can easily do on our own, or with a simple calculator. We don't need R to add two numbers. But where programming languages start to get really powerful is when you have to do the same thing over and over. Today, we're going to learn how R deals with lists of multiple numbers or strings. These lists are called *vectors*. (Actually, a 'list' is also a thing in R, but it means something slightly different... very confusing. So let's stick with calling these things vectors.)

We can create a vector by using `c()`, with the things we want to put inside the vector going in parentheses. (c stands for concatenate/combine.) For example:
```{r}
# Create a vector
vector_1 <- c(1, 4, 3, 18)
print(vector_1)
```

You can use variables to define vectors too! Try it below:
```{r}
# Create 3 variables, named variable_1, variable_2, and bob, and have each of
# them hold a single number

# Combine these variables in any order you want into a vector named bob_the_vector

# print out bob_the_vector

```

Vectors can also hold character types!
```{r}
# Create a vector, char_vector_1 containing multiple strings, and print it out

```

Let's check the *types* that these vectors belong to
```{r}
# check the type of bob_the_vector

# check the type of char_vector_1

# by the way... what is the type of c, as in, the thing you put before your vector?

```

What if we wanted to store the number 3 and the character "3" together in a vector?
```{r}
# Create a vector, mix_vector, with 3 and "3"

# check the type of mix_vector

# why is this happening (check the documentation of c)
```


# Making consecutive sequences

One really useful thing we can do in R is make vectors of evenly spaced numbers. To do this, we can use the `seq()` function. Let's try to use the built-in R documentation of `seq()` to figure out how to use it.
```{r}
# On your own:
# Based on the documentation from of seq, make the sequence: 1, 2, 3, 4

# In pairs:
# Now make the sequence: 2, 2.5, 3, 3.5

# Identify what values are corresponding to what arguments

```

Seq can work in both directions (i.e., the numbers don't have to be going up)
```{r}
# Use seq to make the following sequence: 5, 3, 1, -1

```

There are three things that make seq really powerful compared to just typing out vectors of numbers by hand:
1. It can generate reaaaaaally long sequences of numbers
2. Like all functions, it can use variables as inputs, which means you don't have to change your whole code every time you change your mind about what numbers you're interested in working with
3. When you want to make a vector of a certain length, it will do the math for you of what the spacing between your numbers needs to look like

***
*Skip (all or part) if behind*
Let's try using some of this!
```{r}
# Make a variable number_elements, which will tell you how many numbers you want
# in your final vector

# Make a variable first_number, containing any number you want your vector to
# start with

# Make a variable last_number, containing any number you want your vector to
# end with

# Use first_number, last_number, and number_elements as arguments to seq to
# create your vector, final_vector

```
***

Relatedly, a really useful function in R is `length()`. It can tell you how long your vector is (how many elements are in your vector)
```{r}
# Use length() to find out how long number_elements is

# Use length() to find out how long final_vector is

# Does this make sense?

```

# Operations on vectors of numbers

One of the best things about vectors that will come up again and again soon is that you can do operations on them, just like with numbers. Here's one example:
```{r}
# Create a variable holding a vector that contains an evenly spaced sequence of
# five numbers, from 2 to 4

# Add 3 to the vector. What happens?

```

You can also add vectors together!
```{r}
vector_a <- c(1,2,3)
vector_b <- c(4,5,6)

# Add vector_a and vector_b
```

And finally... many math-related functions can work on vectors! `sum()` is a great example of this.
```{r}
# Create a vector of consecutive numbers between 1 and 100

# Use sum() to add up those numbers

```

# Things we hope you've learned today (and will hopefully remember next time)

* How to look at your **Environment**
* How to use functions, and what *arguments* are
* How to use `sum()`, `paste()`, `seq()`
* Vectors are lists of either numeric or character type
* You can do math on numeric vectors


