---
title: "NYU SURP 2019 R course Class 6: Logicals (continued) Introduction to data frames"
author:
  - Eugene Plavskin
  - Grace Avecilla
  - Tobi Shraink
date: "07/11/2019"
output:
  html_notebook:
    code_folding: show
    depth: 3
    tidy: yes
    toc: yes
---

# Review

```{r}
review_vector <- c(1, 2.6, NA, 4.98, 1.5, NA)

# run a function that will determine which of the values in review_vector is a
# missing value
is.na(review_vector)
# run an operation that will determine which of the numbers in review_vector is
# greater than 2
num_vec <- review_vector > 2
print(num_vec)
```

# Logicals (continued)

## Treating numbers as logicals

In R, FALSE is also encoded as 0, and TRUE is encoded as 1. 

```{r}
0==FALSE
1==TRUE

review_vector_2 <- c(T, T, F, F, T, F)

# How do you think we could easily figure out how many TRUEs there are in
# review_vector_2?
sum(review_vector_2)

# How can you get R to tell you the number of missing values in review_vector?
na_vec <- is.na(review_vector)
sum(na_vec)
sum(is.na(review_vector))
```

## Exercise

Imagine that you're doing an experiment measuring plant height, but some of the plants didn't grow. You decide those should be considered missing values, so you entered `NA` for those plants. Let's count the number of plants that DID grow in your experiment.

```{r}
plant_heights <- c(1.9, 0.1, NA, 0.8, 0.4, 0.2, 7.9, NA, 16.8, 3.5)

# create a logical vector, plant_NA_vector, that contains information on whether
# each value in plant_heights is missing (NA)
plant_NA_vector <- is.na(plant_heights)

# use plant_NA_vector to figure out how many plants total didn't grow
sum(is.na(plant_heights))

# use plant_NA_vector to figure out how many plants DID grow
print(plant_NA_vector)
# one way
sum(plant_NA_vector == FALSE)
# a simpler way
sum(!plant_NA_vector) # remember that ! means 'not'

# bonus:figure out what PROPORTION of plants grew
sum(!plant_NA_vector) / length(plant_NA_vector)
```

# Selecting and replacing values from vectors based on logicals

In addition to using indices to get R to tell us a specific position in a vector, we can also use them to replace specific values. Take a look at this example.
```{r}
vector_to_sub <- c(12, 13, 14, 15, NA, 17, 18, 19)
vector_to_sub[2]
# replace the 2nd value in vector_to_sub with 54
vector_to_sub[2] <- 54
print(vector_to_sub)

# now, try replacing the 1st, 3rd, and 8th values in vector_to_sub with 28,
# using only one line of code (i.e. do it all at once; hint: use a vector)
vector_to_sub[c(1,3,8)] <- 28
print(vector_to_sub)

```

One really powerful way to use this vector substitution is to combine it with logical vectors. Take a look at the example below:
```{r}
vector_to_sub2 <- 1:3
print(vector_to_sub2)
vector_to_sub2[c(TRUE, FALSE, TRUE)] <- 42
print(vector_to_sub2)
# What do you think vector_to_sub2 looks like now? Print it out to check

```

Try this for yourself. Lots of times, we want to replace missing values in vectors. Let's say that, when you do your analysis on the plant example above, you want to treat the plants that didn't grow as having a height of 0 rather than being missing.
```{r}
# use plant_NA_vector to replace every missing value in plant_heights with a 0
plant_heights[plant_NA_vector] <- 0
# print the modified plant_heights vector
plant_heights
```
